---
title: "spatialRF: easy spatial regression with Random Forest"
output:
  github_document:
    toc: true
    toc_depth: 2
    pandoc_args: --webtex
always_allow_html: yes
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

## Introduction

The package **spatialRF** facilitates fitting spatial regression models on regular or irregular data with Random Forest, using the **ranger** package under the hood [(Wright and Ziegler 2017)](file:///tmp/mozilla_blas0/v77i01.pdf), and does so by generating *spatial predictors* that allow the model to take into account the spatial structure of the training data. The end goal is minimizing the spatial autocorrelation of the model residuals as much as possible. Spatial autocorrelation in the residuals indicate that there is a source of spatial autocorrelation in the model variance that cannot be explained by the model predictors, and it is a clear sign that there are important variables missing from the model. In an ideal model, the residuals are not autocorrelated, and should be centered around zero, but this is rarely the case when working with spatial data.

This package implements two main methods to generate *spatial predictors* from the distance matrix of the data points:


  + Principal coordinate analysis of neighbor matrices [(Dray, Legendre, and Peres-Neto 2006)](https://www.sciencedirect.com/science/article/abs/pii/S0304380006000925).
  + Distance matrix columns as explanatory variables [(Hengl et al. 2018)](https://peerj.com/articles/5518/).

The package also provides a set of tools to identify variable interactions, tune random forest hyperparameters, assess model performance on spatially independent data folds, and examine the resulting models via importance plots, and response curves and surfaces. 

## Installing the package

The package is not yet in the CRAN repositories, so at the moment it must be installed from GitHub as follows.

```{r}
remotes::install_github(
  repo = "blasbenito/spatialRF", 
  ref = "main",
  quiet = TRUE
  )
library(spatialRF, quietly = TRUE)
```

There are a few other libraries that will be useful during this tutorial.

```{r}
library(kableExtra)
library(ggplot2)
library(dplyr)
options(dplyr.summarise.inform = FALSE)
library(patchwork)
library(rnaturalearth)
library(rnaturalearthdata)
```


## Fitting spatial models with `spatialRF`

Let's start with a simple example, to then unpack what is happening inside the package functions.

### The data

The package includes an example dataset named **plant_richness_df**, a data frame with plant species richness and predictor variables for 227 ecoregions in the Americas;  a distance matrix among the ecoregion edges named, well, **distance_matrix**, and **plant_richness_sf**, an sf file containing the centroids of the polygons represented in **plant_richness_df**.

```{r}
data(plant_richness_df)
help(plant_richness_df)
data(plant_richness_sf)
data(distance_matrix)
```

The response variable of **plant_richness_df** is "richness_species_vascular", with the total count of vascular plant species found on each ecoregion. The figure below shows the centroids of each ecoregion along with their associated value of the response variable.

```{r, echo=FALSE, fig.width=6, fig.height=5.5}
pr <- cbind(plant_richness_df, plant_richness_sf)
world <- rnaturalearth::ne_countries(
  scale = "medium", 
  returnclass = "sf"
  )

ggplot2::ggplot() +
  ggplot2::geom_sf(data = world, fill = "white") +
  ggplot2::geom_sf(data = pr,
          ggplot2::aes(
            geometry = geom_centroids,
            color = richness_species_vascular
            ),
          size = 2.5
          ) +
  ggplot2::scale_color_viridis_c(direction = -1) +
  ggplot2::theme_bw() +
  ggplot2::labs(color = "Plant richness") +
  ggplot2::scale_x_continuous(limits = c(-170, -30)) +
  ggplot2::scale_y_continuous(limits = c(-58, 75))  +
  ggplot2::ggtitle("Plant richness of the American ecoregions")
```

The predictors (columns 5 to 21) represent diverse factors such as sampling bias, climatic variables, human presence and impact, topography, geographical fragmentation, and features of the neighbors of each ecoregion. The figure below shows the scatterplots of the response variable (y axis) against each predictor (x axis).

```{r, echo = FALSE, fig.width=14, fig.height=9}
plot.list <- list()
for(variable in colnames(plant_richness_df)[5:21]){
  plot.list[[variable]] <- ggplot2::ggplot(
    data = plant_richness_df,
    ggplot2::aes_string(
      x = variable,
      y = "richness_species_vascular",
      color = "richness_species_vascular"
    )
    ) +
    ggplot2::geom_point() +
    ggplot2::scale_color_viridis_c(direction = -1) + 
    ggplot2::theme_bw() + 
    ggplot2::theme(legend.position = "none")
}
patchwork::wrap_plots(plot.list)
```

### Fitting a non-spatial model

To fit a Random Forest model we use the `rf()` function, that takes the data, the names of the response and the predictors, the distance matrix, and a vector of distance thresholds (in the same units as the distances in **distance_matrix**). These distance thresholds are the neighborhoods at which the model will check the spatial autocorrelation of the residuals. Their values may depend on the spatial scale of the data, and the ecological system under study.

```{r}
model.non.spatial <- rf(
  data = plant_richness_df,
  dependent.variable.name = "richness_species_vascular",
  predictor.variable.names = colnames(plant_richness_df)[5:21],
  distance.matrix = distance_matrix,
  distance.thresholds = c(0, 1500, 3000),
  seed = 100,
  verbose = FALSE
)
```

The spatial autocorrelation of the residuals can be plotted with `plot_moran()`

```{r, fig.width=6, fig.height=3, message=FALSE, warning=FALSE}
plot_moran(model.non.spatial)
```

According to the plot, the spatial autocorrelation of the residuals is highly positive for the distances 0 and 1500, while it becomes non-significant (p-value > 0.05, whatever that means) at 3000km. This model is definitely missing something.

### Fitting a spatial model

The non-spatial model fitted above can be converted into a spatial model easily with `rf_spatial()` (this is because **model.non.spatial** has a slot named "ranger.arguments" that contains the model configuration and the data used to train it).

```{r}
model.spatial <- spatialRF::rf_spatial(
  model = model.non.spatial,
  method = "mem.moran.sequential", #default method
  verbose = FALSE
  )
```

The plot below compares the Moran's I of the residuals of the spatial (green) and non spatial (purple) models. It shows that `rf_spatial()` has managed to reduce the spatial autocorrelation (measured via [Moran's I](https://en.wikipedia.org/wiki/Moran%27s_I) with the function `moran_multithreshold()`) of the model residuals for the three selected distances.

```{r, fig.width=6, fig.height=3, message=FALSE, warning=FALSE}
plot_moran(model.spatial)
```

It has done so by generating spatial predictors from the distance matrix, and introducing them into the model one by one until the spatial autocorrelation of the residuals is reduced as much as possible. If we compare the variable importance plots of both models, we can see that the spatial model has an additional set of dots under the name "spatial_predictors", and that the maximum importance of a few of these spatial predictors matches the importance of the most relevant non-spatial predictors.

```{r, fig.width=9, fig.height=6}
p1 <- plot_importance(
  model.non.spatial, 
  verbose = FALSE) + 
  ggplot2::ggtitle("Non-spatial model") 
p2 <- plot_importance(
  model.spatial,
  verbose = FALSE) + 
  ggplot2::ggtitle("Spatial model")
p1 | p2 
```
If we take a look to the five most important variables in **model.spatial** we will see that a few of them are spatial predictors.

```{r, echo = FALSE}
kableExtra::kable(
  head(model.spatial$variable.importance$per.variable)
) %>% kableExtra::kable_styling()
```

Spatial predictors, as shown below, are smooth surfaces representing neighborhood among records at different spatial scales. But where do they come from? How are they generated and included in the model?

```{r, echo=FALSE, fig.width=9, fig.height=5.5}
spatial.predictors <- get_spatial_predictors(model.spatial)
pr <- cbind(spatial.predictors, plant_richness_sf)

p1 <- ggplot2::ggplot() +
  ggplot2::geom_sf(data = world, fill = "white") +
  ggplot2::geom_sf(data = pr,
          ggplot2::aes(
            geometry = geom_centroids,
            color = spatial_predictor_0_6
            ),
          size = 2.5
          ) +
  ggplot2::scale_color_viridis_c(direction = -1) +
  ggplot2::theme_bw() +
  ggplot2::labs(color = "Eigenvalue") +
  ggplot2::scale_x_continuous(limits = c(-170, -30)) +
  ggplot2::scale_y_continuous(limits = c(-58, 75))  +
  ggplot2::ggtitle("Spatial predictor 'spatial_predictor_0_6'") + 
  ggplot2::theme(legend.position = "bottom")

p2 <- ggplot2::ggplot() +
  ggplot2::geom_sf(data = world) +
  ggplot2::geom_sf(data = pr,
          ggplot2::aes(
            geometry = geom_centroids,
            color = spatial_predictor_1500_2
            ),
          size = 2.5
          ) +
  ggplot2::scale_color_viridis_c(direction = -1) +
  ggplot2::theme_bw() +
  ggplot2::labs(color = "Eigenvalue") +
  ggplot2::scale_x_continuous(limits = c(-170, -30)) +
  ggplot2::scale_y_continuous(limits = c(-58, 75))  +
  ggplot2::ggtitle("Spatial predictor 'spatial_predictor_1500_2'") + 
  ggplot2::theme(legend.position = "bottom")

p1 | p2

```

### Generation and selection of spatial predictors

The final model includes `r length(model.spatial$selection.spatial.predictors$names)` spatial predictors. In this particular model, that uses the method "mem.moran.sequential" (Moran's Eigenvector Maps as described in [(Dray, Legendre, and Peres-Neto 2006)](https://www.sciencedirect.com/science/article/abs/pii/S0304380006000925)), the spatial predictors have been generated by the function `mem_multithreshold()`, for every distance threshold selected by the user, as follows:


  + `weights_from_distance_matrix()` sets the values of the distance matrix below a given distance threshold to 0, computes `1/distance_matrix` to convert distances into weights, sets the diagonal to 0, and normalizes the weights.
  + `double_center_distance_matrix()` double-centers the distance matrix, so the sum of each row and each column is zero.
  + `mem()` computes the positive eigenvectors of the double-centered matrix.

The function `rank_spatial_predictors()` ranks the spatial predictors from higher to lower spatial autocorrelation (computed with `moran()`), and `select_spatial_predictors_sequential()` introduces them one by one into the model, and selects the smallest subset that minimizes the Moran's I of the residuals while maximizing the R squared of the model, as shown in the optimization plot below (every dot represents one spatial predictor, and its size matches their position in the ranking, and their order of inclusioin in the model).

```{r, echo=FALSE, fig.width=6, fig.height=4}
p <- plot_optimization(model.spatial)
```
## Assessing model performance

Models fitted with `rf()` or `rf_spatial()` have a performance slot with several performance measures.

```{r}
print_performance(model.spatial)
```

R squared is computed on the out-of-bag data (fraction of data not used while training each regression tree in the forest), while the other three performance measures are computed by comparing observations and model predictions. These performance scores are not computed on independent data, and therefore do not represent the model ability to predict over *unseen* data.

The function `rf_evaluate()` separates the data into a number of spatially independent training and testing folds, fits a model on each training fold, predicts over each testing fold, and computes performance measures, to finally aggregate them across model repetitions. Let's see how it works.

```{r}
model.spatial <- rf_evaluate(
  model = model.spatial,
  xy = plant_richness_df[, c("x", "y")], #data coordinates
  repetitions = 30,                      #number of folds
  training.fraction = 0.75,              #training data fraction
  verbose = FALSE
)
```

The function generates a new slot in the model named "evaluation" with several objects that summarize the spatial cross-validation results.

```{r}
names(model.spatial$evaluation)
```


```{r, fig.width=6, fig.height=6}
plot_evaluation(model.spatial)
```




